Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-23 draft.Rmd
uploaded working model

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

@@

This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan Lapagog lapagos in Northeastern Scandinavia (). Although not in immediate danger, the willow ptarmigan faces threats from habitat loss, climate change, and hunting and trapping. Hjeljord and Loe postulate that, along with climate, a long-term dampening of the amplitude in small rodents and an increase in red fox numbers, have prevented willow ptarmigan populations from reaching their peaks seen a hundred years ago (). Their analysis implements linear models with a count proxy as a function of time to estimate linear change and wavelet analysis to detect cyclic periods. This report’s aim is to further Hjelford and Loe’s work by capturing the stochastic population dynamics and the role of alternative prey using the partially observed markov process (POMP) framework

@@

Surprisingly, grouse, the family of birds which willow ptarmigan belong to, are of the most studied bird species in the world. Thus, an abundance of scientific research has been carried out to shed light on the decline in its population. @pedersen found that intensified forestry practices have significantly reduced suitable breeding habitat. Additionally, agricultural expansion has led to the conversion of tundra and heathland habitats, further diminishing ptarmigan populations [@henden]. Climate change exacerbates these effects by altering snow cover patterns, impacting the ptarmigans' ability to camouflage and increasing predation risk. Research by @tottrup highlighted that warmer temperatures and reduced snow cover are leading to increased predation pressure on ptarmigans in Scandinavia. Furthermore, hunting pressure has also played a role in the decline of ptarmigan populations. Studies by @hebbelstrup and @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.  Breisjøberget el al., found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis.

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
```

```{r data_table, echo=FALSE}
data <- read_excel("./data/bird_data.xlsx")
data |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

knitr::kable(data[subset], caption = "Data")
```

@@

p2 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

p3 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  geom_rect(
    data = subset(data, peak_rodent_year == 1),
            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
            fill = "blue", alpha = 0.3
    ) + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

@@

```{r, boxplot, echo=FALSE}
ggplot(data, aes(x = peak_rodent_year, y = log.CPUE)) +

  geom_boxplot() + 
  theme_light() + 
  xlab("Peak Rodent Year")
```

@@

### ACF
```{r ACF, echo=FALSE}
acf(data$log.CPUE, lag = 100, main = "ACF log.CPUE")
```

@@

```{r PACF, echo=FALSE}
pacf(data$log.CPUE, lag = 100, main = "PACF log.CPUE")
```
## {-}
--------

@@

bird_aic_table <- aic_table(data$log.CPUE,4,5)
require(knitr)
kable(bird_aic_table,digits=2)
```

@@

```{r}
statenames <- c("logF", "logB")
paramnames <- c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs")
obs_names <- c("logCPUE")
covarnames <- c("R")
# logF_0 = 0 normalizing to assume unit initial population
# sigmaF roughly the noise in a year
param_guess <- c(
  logF_0 = 0, 
  logB_0 = 0,
  a = 1,
  b = 2, 
  c = 1,
  alpha = 1, 
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1, 
  sigmaB = 0.1, 
  logRho = 3, 
  sigma_obs = 0.1
)
```

@@

```{r, eval=FALSE}
Csnippet("
  double dwW, dwB;

@@

Csnippet("
  log.CPUE = rnorm(logB + logrho, signma_obs)
") -> rmeas

Csnippet("
  lik = dnorm(logCPUE, lobB + logrho, sigma_obs.give_log)
") -> dmeas

bird |>
  pomp(times="year", t0=1872,
       rprocess=euler(step, delta.t=1/52),
       rinit=rinit,
       rmeasure=rmeas,
       dmeasure=dmeas,
       statenames=statenames,
       paramnames=paramnames,
       obs_names <- c("logCPUE"),
       covarnames <- c("R")
    ) -> mod

```

- `rinit`: simulator for the initial-state distribution, i.e., the distribution of the latent state at time $t_0$

>>>>

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
bibliography: references.bib 
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

@@

This report extends research by Hjeljord and Loe, who explain the dwindling numbers of willow ptarmigan *Lapagos lapagos* in Northeastern Scandinavia (). Although not in immediate danger, the willow ptarmigan faces threats from habitat loss, climate change, and hunting and trapping. Hjeljord and Loe postulate that, along with climate, a long-term dampening of the amplitude in small rodents and an increase in red fox numbers, have prevented willow ptarmigan populations from reaching their peaks seen a hundred years ago (). Their analysis implements linear models with a count proxy as a function of time to estimate linear change and wavelet analysis to detect cyclic periods. This report’s aim is to further Hjelford and Loe’s work by capturing the stochastic population dynamics and the role of alternative prey using the partially observed markov process (POMP) framework

@@

Surprisingly, grouse, the family of birds which willow ptarmigan belong to, are of the most studied bird species in the world. Thus, an abundance of scientific research has been carried out to shed light on the decline in its population. Pedersen el al., @pedersen found that intensified forestry practices have significantly reduced suitable breeding habitat. Additionally, agricultural expansion has led to the conversion of tundra and heathland habitats, further diminishing ptarmigan populations [@henden]. Climate change exacerbates these effects by altering snow cover patterns, impacting the ptarmigans' ability to camouflage and increasing predation risk. Research by Tottrup et al., @tottrup highlighted that warmer temperatures and reduced snow cover are leading to increased predation pressure on ptarmigans in Scandinavia. Furthermore, hunting pressure has also played a role in the decline of ptarmigan populations. Studies by Hebbelstrup @hebbelstrup and Storass @storass documented the impact of hunting on ptarmigan populations in Norway and Sweden, respectively.  Breisjøberget el al., found that red foxes negatively affect ptarmigan when small rodent abundance is low, which is in accordance with the alternative prey hypothesis. Despite there being a plethora of work conducted in this area, none have applied

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(doParallel)
library(pomp)
```

```{r data_table, echo=FALSE}
df <- read_excel("./data/bird_data.xlsx")
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

knitr::kable(head(data[subset], 10), caption = "Data")
```

@@

p2 <- ggplot(data, aes(x=year, y=logCPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

p3 <- ggplot(data, aes(x=year, y=logCPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  geom_rect(
    data = subset(data, R == 1),
            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
            fill = "blue", alpha = 0.3
    ) + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

@@

```{r, boxplot, echo=FALSE}
ggplot(data, aes(x = factor(R), y = logCPUE, group = factor(R))) + 
  geom_boxplot() + 
  theme_light() + 
  xlab("Peak Rodent Year") +
  scale_x_discrete(labels = c("no" = "0", "yes" = "1"))
```

@@

### ACF
```{r ACF, echo=FALSE}
acf(data$logCPUE, lag = 100, main = "ACF log.CPUE")
```

@@

```{r PACF, echo=FALSE}
pacf(data$logCPUE, lag = 100, main = "PACF log.CPUE")
```
## {-}
--------

@@

bird_aic_table <- aic_table(data$logCPUE,4,5)
require(knitr)
kable(bird_aic_table,digits=2)
```

@@

```{r, echo=FALSE, eval=FALSE}
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1))  |>
  rename(reports = "CPUE") |>
  select(c("year", "reports", "R")) -> bird
```

```{r echo=FALSE, eval=FALSE}
birdMod |>
  pfilter(Np=5, params = params) -> pf

mif2(
  measSIR, 
  Nmif = 10, 
  Np = 2000,
  cooling.fraction.50 = 0.5, 
  rw.sd = rw_sd(
    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
  ),
  params = params
) -> mif_out
```

### Local Search

```{r}
statenames <- c("logF", "logB")
paramnames <- c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs")
obs_names <- c("logCPUE")
# logF_0 = 0 normalizing to assume unit initial population
# sigmaF roughly the noise in a year
```

@@

```{r}
Csnippet("
  double dwF, dwB;

@@

Csnippet("
  logCPUE = rnorm(logB - logRho, sigma_obs);
") -> rmeas

Csnippet("
  lik = dnorm(logCPUE, logB - logRho, sigma_obs, give_log);
") -> dmeas
```

```{r}
bird1 <- bird |> 
  dplyr::select(-R)

bird1 |>
  pomp(
    times="year", 
    t0=1871,
    rprocess=euler(step, delta.t=1/52),
    rinit=rinit,
    rmeasure=rmeas,
    dmeasure=dmeas,
    statenames=statenames,
    paramnames=paramnames,
    obs_names <- c("logCPUE"),
    covar = covariate_table(
      bird |> dplyr::select(year, R) |> 
        dplyr::bind_rows(c("year" = 1871, R = 0)) |> 
        dplyr::arrange(year), 
      times = 'year',
      order = "constant"
    ),
    partrans = parameter_trans(
      log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
              'c', 'logF_0', 'logB_0', 'logRho')
    )
    # covarnames <- c("R")
  ) -> mod
```

```{r}
param_guess <- c(
  logF_0 = 1, 
  logB_0 = 1,
  a = 1,
  b = 2, 
  c = 1,
  alpha = 1, 
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1, 
  sigmaB = 0.1, 
  logRho = 3, 
  sigma_obs = 0.1
)
```

```{r}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out
```

```{r}
freeze(
  runif_design(
    lower = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0)
    upper = c(alpha = 0, Beta = 0, gamma = 0, a = 0, b = 0, c = 0,
              sigma_obs = 0, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
              sigmaF = 0, sigmaB = 0.02, logRho = 0)
  ),
  seed = 420
) -> guesses
```

### Global Search
